Defining properties regulating pioneer-factor binding and function

### 2022-12-22 manuscript/manuscript.qmd
update MS results

To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq).
We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh.
To control for antibody specificity, we performed additional ChIP experiments in WT cells using IgG, anti-Zld, or anti-Grh (Extended Data Fig. XX).
We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh.
<!--# Not sure the best way to discuss the control ChIP experiments. Should I mention how many peaks we detect in WT cells? What supplemental figures should I include? Should peak calling be done using IgG or WT cells as control instead of input? -->

@@

Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation.
To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression.
Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig. XX).
Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig. XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression.
<!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing. Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig. XX).
<!--# Add more detailed information on GO analysis here -->

@@

-   For all three factors, increased concentration results in increased binding and opening of class II and III sites.

-   For all three factors, increased concentration leads to novel binding sites.

-   For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.

-   For Grh, increased concentration does allow binding and opening at repressed sites

Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022].
Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering.

@@

Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
However, little is known about how regions outside the DBD may contribute to pioneering function.
<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD -->.
To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. XX).
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig. 6a,c).
We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.

@@

| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
|--------------------|------------------|------------------|------------------|
| pMT-puro                              | AddGene                   | 17923                |                         |
| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
| PVDF Membrane                         |                           |                      |                         |
| Formaldehyde                          |                           |                      |                         |
| Protease inhibitor tablets            |                           |                      |                         |
| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
| TruSeq v2 RNA library prep kit        |                           |                      |                         |
| Trizol                                |                           |                      |                         |
| **Antibodies**                        |                           |                      |                         |
| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB: 1:1000, CR: 0.5 µL  |
|                                       |                           |                      |                         |
|                                       |                           |                      |                         |

@@

**Figure 1: Ectopic pioneer factors open chromatin and activate transcription.**

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites.**

**Figure 3: Motif content shapes pioneer-factor activity**

**Figure 4: H3K27me3 and H3K9me3 are barriers to pioneer-factor binding.**

**Figure 5: Binding and opening of closed chromatin is concentration dependent.**

**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function.**

>>>>

To determine how exogenous Zld and Grh interact with preexisting chromatin in S2 cells, we expressed each factor at physiological levels and determined genome-wide binding sites using chromatin immunoprecipitation coupled with sequencing (ChIP-seq). We identified `r zld_n_ChIP_peaks` peaks for Zld and `r grh_n_ChIP_peaks` for Grh. To control for antibody specificity, we performed additional ChIP experiments in WT cells using anti-Zld, or anti-Grh (Extended Data Fig. XX). We excluded any peaks that were called in these datasets, ensuring that we analyzed peaks that were detected only upon ectopic expression of Zld or Grh. <!--# Not sure the best way to discuss the control ChIP experiments. Should I mention how many peaks we detect in WT cells? What supplemental figures should I include? Should peak calling be done using IgG or WT cells as control instead of input? -->

@@

zld_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Zld_RNAseq_S2-Zld-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

grh_RNAseq_results <- here("RNAseq/results/DEseq2/S2-Grh_RNAseq_S2-Grh-vs-S2-WT_results.tsv") |> 
  read_tsv() |> 
  mutate(direction_of_change = case_when(
    is_diff & log2FoldChange > 0 ~ "increased",
    is_diff & log2FoldChange < 0 ~ "decreased",
    !is_diff ~ "ns"
  ))

zld_diff <- zld_RNAseq_results |> 
  pull(is_diff) |> 
  sum()
zld_up <- (zld_RNAseq_results$direction_of_change == "increased") |> 
  sum()
zld_down <- (zld_RNAseq_results$direction_of_change == "decreased") |> 
  sum()

grh_diff <- grh_RNAseq_results |> 
  pull(is_diff) |> 
  sum()
grh_up <- (grh_RNAseq_results$direction_of_change == "increased") |> 
  sum()
grh_down <- (grh_RNAseq_results$direction_of_change == "decreased") |> 
  sum()

Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation. To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression. Differential expression analysis using DESeq2 [@love2014] identified `r zld_diff` differentially expressed genes for Zld and `r grh_diff` for Grh (Extended Data Fig. XX). Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig. XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression. <!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing. Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig. XX). <!--# Add more detailed information on GO analysis here -->

@@

-   ~~For all three factors, increased concentration results in increased binding and opening of class II and III sites.~~

-   ~~For all three factors, increased concentration leads to novel binding sites.~~

-   ~~For Zld and Twi, increased concentration is insufficient to allow binding to repressed sites.~~

-   ~~For Grh, increased concentration does allow binding and opening at repressed sites~~

Previous studies have proposed that pioneer factor activity may depend on protein concentration [@Hansen2022; @Blassberg2022]. Considering that tazemetostat treatment had limited effects on pioneer factor binding, we asked if varying the protein concentration of Zld, Grh, or Twi could affect binding and pioneering. We induced cells to express Zld, Grh and Twi at a range of protein concentrations: uninduced, subphysiological, physiological, or supraphysiological (Extended Data Fig. XX; see methods). We performed ChIP-seq and analyzed the data at our previously defined class I-VI binding sites. We found that binding and was concentration dependent (Fig. 5). At class II and III binding sites, subphysiological levels of Zld, Grh, and Twi were sufficient for some, albeit reduced binding (Fig. 5a,e,i). Expression of supraphysiological protein levels resulted in increased ChIP intensity at class II and III sites (Fig. 5a,e,i). We also performed ATAC-seq to determine if concentration-dependent increases in binding led to changes in chromatin accessibility. Changes in chromatin accessibility were also concentration dependent, with more limited chromatin opening at lower concentrations and more robust opening at higher concentrations (Fig. 5b,f,j). Intriguingly, many class II binding sites became accessible when Zld, Grh, and Twi were expressed at the highest levels. For Grh, but not for Zld or Twi, supraphysiological concentrations allowed some binding and opening to class IV-VI sites (Fig. 5, Extended Data Fig. XX). To determine how protein concentration changed the distribution of class I, II and III sites, we redefined these classes using the ChIP-seq and ATAC-seq data at each concentration. For Zld and Twi, supraphysiological concentrations led to an increase in class III regions relative to class II regions (Fig. 5c,k). For Grh, higher concentrations caused a dramatic increase in binding to closed chromatin, with the majority of these sites being falling in III (Fig. 5g). To assess the extent to which increased concentrations led to binding to new regions that were not bound at physiological concentrations, we overlapped the binding sites detected at physiological vs. supraphysiological concentrations (Fig. 5d,h,i). For all three proteins, there were novel binding sites that were bound only when expressed at the highest concentration. Together, these data demonstrate that both binding and chromatin opening by pioneer factors are concentration dependent, and that increasing protein concentration leads to novel binding sites, including some sites that are restistant to binding at physiological levls.

@@

Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs). However, little is known about how regions outside the DBD may contribute to pioneering function. <!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD. See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->. To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. XX). Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig. 6a,c). We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a]. Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding. To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif. This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required. Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e). For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f). Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.

@@

| Reagent                               | Source/vendor             | Catalog/accession \# | Notes                   |
|-------------------|------------------|------------------|------------------|
| pMT-puro                              | AddGene                   | 17923                |                         |
| Gibson Cloning                        | New England Biolabs       | E2611                |                         |
| Schneider's Medium                    | Thermo Fisher Scientific  | 21720001             |                         |
| Fetal Bovine Serum                    | Omega Scientific          | FB-02                |                         |
| Antibiotic/antimycotic                | Thermo Fisher Scientific  | 15240062             |                         |
| Effectene Transfection Reagent        | Qiagen                    | 301427               |                         |
| PVDF Membrane                         |                           |                      |                         |
| Formaldehyde                          |                           |                      |                         |
| Protease inhibitor tablets            |                           |                      |                         |
| NEBNext Ultra II DNA library prep kit |                           |                      |                         |
| TruSeq v2 RNA library prep kit        |                           |                      |                         |
| Trizol                                |                           |                      |                         |
| **Antibodies**                        |                           |                      |                         |
| Anti-Zld                              | Harrison et al., 2010     |                      | ChIP: 6 µL , WB: 1:750  |
| Anti-Grh                              | Harrison et al., 2010     |                      | ChIP: 8 µL, WB: 1:1000  |
| Anti-Twi                              | He et al., 2015           |                      | ChIP: 10 µg, WB: 1:1000 |
| Anti-HA-peroxidase clone 3F10         | Roche                     | 12013819001          | WB: 1:500               |
| Anti-Tubulin DM1A                     | Sigma                     | T6199-200UL          | WB: 1:5000              |
| goat anti-rabbit IgG-HRP conjugate    | Bio-Rad                   | 1706515              | WB: 1:3000              |
| Anti-HA clone 12CA5                   | Roche                     |                      | ChIP: 7.5 µL            |
| Anti-H3K27me3                         | Cell Signaling Technology |                      | WB: 1:1000, CR: 0.5 µL  |
|                                       |                           |                      |                         |
|                                       |                           |                      |                         |

@@

**Figure 1: Ectopic pioneer factors open chromatin and activate transcription. a-b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**. ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for WT cells lacking Zld or Grh expression. **c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**. **d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**. **e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h). i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions where changes in chromatin accessibility are associated with changes in gene expression for Zld **(i)** or Grh **(k)**. **j,i,** Proportion of class I, II or III regions that are proximal to a gene that is upregulated upon Zld **(j)** or Grh **(i)** expression. Promoter-proximal and promoter-distal binding sites are shown separately. The grey dotted line indicates the percentage of all binding sites that are proximal to a upregulated expressed gene.

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites. a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites. **c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions. **d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.

**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi. **b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi. Zld, Grh, and Twi ChIP-seq data were generated in this study. Remaining ChIP-seq datasets were previously published (see Supplementary table XX). **c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**. Logos plot of the canonical motif are shown alongside the heatmap. Percentages are shown for class I, II or III binding sites, or all WT ATAC-seq peaks as a control for the background motif frequency. **f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**. **g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**. Motif scores were determined by comparing each identified motif instance to the canonical motif.

**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**. ChIP-seq data for repressive histone modifications H3K27me3 and H3K9me3 is also shown. **d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**. Zld, Grh, and Twi ChIP-seq data are z-score normalized. CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).

**Figure 5: Binding and opening of closed chromatin is concentration dependent. a,e,i,** Metaplots of ChIP-seq signal at class I-VI regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels. Concetrations of CuSO~4~ used to induce protein expression are indicated above each plot. **b,f,j,** Metaplots of ATAC-seq signal at class I-VI regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels. **c,g,k,** Distribution of class I, II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**. Y-axis indicates the total number of peaks called at each concentration. Colors indicate the proportion of those peaks that are defined as class I, II, or III. **d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.

**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function. a-b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening. Shown for Zld **(a)** and Grh **(b)**. **c,g,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(g)**. **d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
